---
title: "Hungarian morphological variation with JANET"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hungarian morphological variation with JANET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

This vignette demonstrates JANET on Hungarian verbal morphology. Some Hungarian
verbs vary between the suffixes *-k* and *-m* in the first-person singular
indefinite (see Rácz & Lukács 2024). The intuition is that a novel verb will
behave like its nearest phonological neighbours in the lexicon — verbs that are
more similar to *-k* verbs will prefer *-k*; those closer to *-m* verbs will
prefer *-m*.

JANET computes the phonological distances that feed such an analogy model. Here
we walk through the full pipeline and inspect the output.

## Setup

```{r load}
library(janet)
```

Read the Hungarian segmental feature matrix and word list from the package's
bundled data files.

```{r read-data}
features_path <- system.file("extdata", "hungarian_features.tsv", package = "janet")
words_path    <- system.file("extdata", "hungarian_words.tsv",    package = "janet")

features <- read.delim(features_path, encoding = "UTF-8", na.strings = "")
words_df  <- read.delim(words_path,   encoding = "UTF-8")
words     <- words_df$lemma
```

The feature matrix covers `r nrow(features)` Hungarian segments and
`r ncol(features) - 1` binary distinctive features.

```{r preview-features}
head(features[, 1:6])
```

The word list contains `r length(words)` verb stems.

```{r preview-words}
head(words)
```

## Natural classes

We can inspect the natural classes without running the full pipeline.

```{r natural-classes}
nc <- generate_natural_classes(features)
nrow(nc)
head(nc, 10)
```

`r nrow(nc)` natural classes are derived from the feature matrix.

## Segment similarity

```{r segment-similarity}
seg_sim <- compute_segment_similarity(features, nc)
nrow(seg_sim)
```

A few illustrative pairs — short vowels versus long counterparts, and voicing
pairs among obstruents:

```{r sim-examples}
pairs <- data.frame(
  s1 = c("a", "e", "p", "t", "k"),
  s2 = c("á", "é", "b", "d", "g")
)
for (i in seq_len(nrow(pairs))) {
  row <- seg_sim[seg_sim$segment1 == pairs$s1[i] & seg_sim$segment2 == pairs$s2[i], ]
  cat(sprintf("sim(%s, %s) = %.3f\n", pairs$s1[i], pairs$s2[i], row$similarity))
}
```

Short/long vowel pairs are highly similar (they differ on only a few features);
voicing pairs among stops are moderately similar.

## Full pipeline on a small subset

Running the full 305-word matrix takes a minute. For illustration we use a
20-word subset.

```{r run-janet}
set.seed(42)
sample_words <- sample(words, 20)
result <- run_janet(features, sample_words, gap_penalty = 1.0)
```

### Word distances

```{r word-distances}
head(result$word_distances[result$word_distances$word1 != result$word_distances$word2, ])
```

### Alignment log

```{r alignment-log}
head(result$alignment_log)
```

The `alignment1` and `alignment2` columns show the optimal segment-by-segment
pairing; `distances` gives the per-position cost; `summed_distance` is the
total phonological distance.

## Nearest neighbours

For each word we can find its closest phonological neighbour.

```{r nearest-neighbours}
library(dplyr)

result$word_distances |>
  filter(word1 != word2) |>
  group_by(word1) |>
  slice_min(phon_dist, n = 1L, with_ties = FALSE) |>
  ungroup() |>
  arrange(word1) |>
  head(10)
```

## References

Frisch, S. A., Pierrehumbert, J. B., & Broe, M. B. (2004). Similarity
avoidance and the OCP. *Natural Language & Linguistic Theory*, 22(1), 179–228.

Dawdy-Hesterberg, L. G., & Pierrehumbert, J. B. (2014). Learnability and
generalisation of Arabic broken plural nouns. *Language, Cognition and
Neuroscience*, 29(10), 1268–1282.

Rácz, P., & Lukács, Á. (2024). Variation in the 1sg. indef: More than you
wanted to know. *Acta Linguistica Academica*, 71(1–2), 2–17.
